
---
title: "Rosenbaum Bounds Sensitivity Analysis in Ecology"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates how to perform **Rosenbaum Bounds Sensitivity Analysis** using R, applied to an ecological study assessing the impact of deforestation on bird species richness.

## Load Required Libraries

```{r}
# Install packages if not already installed
if (!require(MatchIt)) install.packages("MatchIt", dependencies = TRUE)
if (!require(rbounds)) install.packages("rbounds", dependencies = TRUE)

# Load the libraries
library(MatchIt)
library(rbounds)
```

## Simulate Ecological Data

We create a dataset where bird species richness is compared between deforested and non-deforested sites while controlling for elevation and precipitation.

```{r}
set.seed(123)
data <- data.frame(
  deforestation = rep(c(1, 0), each = 50),
  species_richness = c(rnorm(50, mean = 10, sd = 3), rnorm(50, mean = 15, sd = 3)),
  elevation = rnorm(100, mean = 500, sd = 100),
  precipitation = rnorm(100, mean = 1000, sd = 200)
)
```

## Perform Propensity Score Matching

We match deforested and non-deforested sites based on elevation and precipitation.

```{r}
match_model <- matchit(deforestation ~ elevation + precipitation, data = data, method = "nearest")
matched_data <- match.data(match_model)
```

## Conduct Sensitivity Analysis with Rosenbaum Bounds

We check how sensitive our results are to hidden bias.

```{r}
# Extract matched treatment and control outcomes
treated_outcomes <- matched_data$species_richness[matched_data$deforestation == 1]
control_outcomes <- matched_data$species_richness[matched_data$deforestation == 0]

# Perform Rosenbaum sensitivity analysis
psens(x = treated_outcomes, y = control_outcomes, Gamma = 1.5)  # Adjust Gamma to test sensitivity
```

## Interpretation of Results

- If the p-value remains small at **Γ = 1.5**, our conclusions are robust to moderate hidden bias.
- If the p-value becomes insignificant at lower values of **Γ**, our study is highly sensitive to hidden confounders.

## Conclusion

Rosenbaum bounds allow us to assess the robustness of our ecological study results against unmeasured confounding. This is particularly useful when controlled experiments are infeasible.

### Take Aways

Rosenbaum bounds provide a way to assess how much an unmeasured confounder could influence the causal relationship between deforestation and bird species richness. In ecological studies, where controlled experiments are often impractical, this method helps gauge the reliability of observational findings.
